---
title: "Segregated Choice: The Impact of Public School Closures on a Community"
author: "Anthony Kuykendall"
date: "December 5, 2018"
output: beamer_presentation
bibiliography: DCPSBib.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(maps)
library(UScensus2010)
library(UScensus2010tract)
library(sp)
library(maptools)
library(rgeos)
library(ggmap)
# Reading and Tidying CSV Data Set
closed_schools <- read.csv("ClosedDCPS.csv")
colnames(closed_schools)[colnames(closed_schools)=="X"] <- "Longitude"
colnames(closed_schools)[colnames(closed_schools)=="Y"] <- "Latitude"
colnames(closed_schools)[colnames(closed_schools)=="FACUSE"] <- "Designation"
colnames(closed_schools)[colnames(closed_schools)=="PS"] <- "PreSchool"
colnames(closed_schools)[colnames(closed_schools)=="PK"] <- "PreK"
colnames(closed_schools)[colnames(closed_schools)=="GR_1"] <- "Grade 1"
colnames(closed_schools)[colnames(closed_schools)=="GR_2"] <- "Grade 2"
colnames(closed_schools)[colnames(closed_schools)=="GR_3"] <- "Grade 3"
colnames(closed_schools)[colnames(closed_schools)=="GR_4"] <- "Grade 4"
colnames(closed_schools)[colnames(closed_schools)=="GR_5"] <- "Grade 5"
colnames(closed_schools)[colnames(closed_schools)=="GR_6"] <- "Grade 6"
colnames(closed_schools)[colnames(closed_schools)=="GR_7"] <- "Grade 7"
colnames(closed_schools)[colnames(closed_schools)=="GR_8"] <- "Grade 8"
colnames(closed_schools)[colnames(closed_schools)=="GR_9"] <- "Grade 9"
colnames(closed_schools)[colnames(closed_schools)=="GR_10"] <- "Grade 10"
colnames(closed_schools)[colnames(closed_schools)=="GR_11"] <- "Grade 11"
colnames(closed_schools)[colnames(closed_schools)=="GR_12"] <- "Grade 12"
colnames(closed_schools)[colnames(closed_schools)=="TOTAL_STUD"] <- "Total Students"

# Snapshot of Schools Closed (Name, Designation, Year)
school_designation <- subset(closed_schools, select = c(4, 6, 39))

# Code for DC Map Figure Overlay with DCPS Closure Locations
data("district_of_columbia.tract10")
n_tract <- length(district_of_columbia.tract10)
DC_base <- ggplot(data = district_of_columbia.tract10) + 
  geom_polygon(mapping = aes(x = long, y = lat), color = "gray",
                                    fill = NA, size = 0.5)
DCPS_closures <- DC_base + 
  geom_point(data = closed_schools, aes(x = Longitude, y = Latitude, color = Designation),
                     na.rm = TRUE)

# Code for Grade Breakdown Slide
student_by_grade <- subset(closed_schools, select = c(10:25))
Grade_Breakdown <- colSums(student_by_grade, na.rm = TRUE, dims = 1)

# Code for Demographics
CensusData <- read.csv("Census_Data.csv")
library(data.table)
setattr(CensusData, "row.names", c("5-9", "10-14", "15-17", "18-19", "Total"))

# Code for Wards/School
Ward_Breakdown <- read.csv("Ward_Breakdown.csv")

Ward_Plot1 <- ggplot(data = Ward_Breakdown) +
  geom_point(mapping = aes(x = Total, y = Schools, color = Ward))
Ward_Plot2 <- ggplot(data = Ward_Breakdown) +
  geom_point(mapping = aes(x = Total, y = Schools, color = Ward)) +
  geom_smooth(mapping = aes(x = Total, y = Schools))

# Code for LRM
SchoolReg <- lm(Total~Schools, data = Ward_Breakdown)
```

## Introduction: The Premise

- DCPS has an equity problem
- More affluent, well off neighborhoods are able to outperform schools in struggling neighborhoods
- 
Between 2008-2013: 42 public schools closed, 28 of them elementary schools

## Education and Childhood Development

- A child's educational development is interlinked with cultural developments (Raikes 2017)
- Educational structure can inforce positive cultural norms (Nugent 2014)
- WV: Students lost time for studying, extracurriculars, and outside play due to new commutes established after school closures (Lee 2017)
- Negative social implications to moving schools

## DC Public School Closures

```{r}
DCPS_closures
```

## DC Ward Demographics: A Snapshot
```{r}
as.tibble(CensusData)
```

## Impact: Breakdown by Grade

```{r}
Grade_Breakdown
```

## The Question

-Are you more likely to have a public school closed in your neighborhood if you are a black child than if you are a white child?

## Wards Impacted by School Closures 2008-2013
```{r}
Ward_Breakdown
```

## Wards Impacted by School Closures
```{r}
Ward_Plot1
```

## Wards Impacted by School Closures
```{r}
Ward_Plot2
```

## Linear Regression Model
- In this model, population of African Americans between the ages of 5 and 19 will be used as a predictor for the number of schools closing in a Ward.

## LRM
```{r}
summary(SchoolReg)
```

## Regression Continued
```{r}
with(Ward_Breakdown, plot(Total, Schools))
abline(SchoolReg)
```

## Policy Options
- Schools close, but the story does not stop there
- The rise of charter schools, particularly in DC could help
- Parent participation in schools is a barrier worth looking into

## Confounding Variables
- Neighborhood Diversity within the Wards
- Charter Schools/Non DCPS Presence
- School Size
- School Density in Areas

## Limitations of the Study
- Missing Data in School Closures
- No Accompanying Data for Schools Open
- Neighborhood Demographics Precision